DATA DRIVEN MODELS FOR APOGEE M DWARFS

### 2018-03-26 paper/draft1/draft1.tex
add derivative plots

\begin{enumerate}
\item[-] Why M-dwarfs are important: most abundant type of star, are planet hosts (Trappist 1, other examples?), have lifespans longer than the age of the universe (billions or up to trillions of years) making them an interesting probe of Milky way populations. \color{red}{Need citations!}
\end{enumerate}

\subsection{M dwarf Challenge}
\begin{enumerate}
\item[-] Various attempts / samples for which people have tried to make substantial catalogs of M-dwarfs with physical labels on them (\citealt{West:2011} is the largest M dwarf catalog of $\sim$ 70k sources... other catalogs? -- \color{red}{Need citations!}\color{black})

\item[-] Why is there a problem labeling M dwarfs? Generally, two kinds of approaches to labeling: (1) model the spectra using atmospheric models (BTSettl, PHOENIX, KURUCZ, MARCS, ATLAS, etc.) find some places in literature where people have tried these on M dwarfs (\color{red}{Need citations... }\color{black} look at Burgasser, Blake...) - however missing opacities, and complexities due to molecules/clouds/chemistry make this difficult (\color{red}{Need citations... }\color{black} look at Allard, Homeier...); (2) model narrow wavelegth region where models are good, look at specific features (\citealt{Rojas-Ayala:2012}...), or reduce resolution (use photometries) -- however we want to use \emph{all} of the data!

@@

@@

@@

\subsection{APOGEE Survey}
\begin{enumerate}
\item[-] Describe the APOGEE data (survey mission: \citealt{Majewski:2015}; target selection: \citealt{Zasowski:2013}; ASPCAP pipeline: \citealt{Perez:2016}).

\item[-] Some M dwarfs in APOGEE proposed as ancilliary targets (Blake/Covey sources), plus an unknown number observed unintentionally \citep{Desphande:2013}.

\item[-] Describe M dwarf studies that have already been conducted in APOGEE (\citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs; \citealt{Souto:2017} -- modeled 2 M dwarfs, determined Teff/logg/metallicity + 13 elemental abundances; \citealt{Desphande:2013} -- modeled rotational and radial velocites for several hundred M dwarfs; \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity)
\end{enumerate}

@@

%%==========================================================================================
\section{Experimental Results} \label{sec:results}
% \begin{enumerate}
% \item[-] The spectral-type Cannon. Show the results on your AAS poster, and also show plots of derivatives wrt type, with maybe blow-ups on interesting features.

% \item[-] The parameter Cannon. Same.
% \end{enumerate}

\subsection{Spectral Type Model}
\begin{figure}[ht]
\plotone{figures/west_spt_self_test.png}
\caption{Spectral type label self test.}
\end{figure}

\begin{figure}[ht]
\plotone{figures/1_Spectral_Sequence.pdf}
\plotone{figures/2_Spectral_Sequence.pdf}
\plotone{figures/3_Spectral_Sequence.pdf}
\caption{ Spectral sequence of dwarfs in training set M0-M9; chip 1 of APOGEE
spectrum with highlighted spectral type sensitive regions identified in \citealt{Desphande:2013}.}
\end{figure}

\subsection{Temperature/Metallicity Model}
\begin{figure}[ht]
\plottwo{figures/mann_teff_self_test.png}{figures/mann_feh_self_test.png}
\caption{Temperature and Metallicity label self test.}
\end{figure}

\begin{figure}[ht]
\plotone{figures/mann_model_demo.png}
\caption{\textit{Top two plots:} Mann-trained model for varying temperatures; \textit{bottom two plots:} Mann-trained model for varying metallicities.}
\end{figure}

@@

\item[-] The Cannon as a test of evaluating training parameter accuracy? How to analyze this?... \color{red}{Maybe add training on theoretical models? Could be a test of seeing how well The Cannon could derive labels if there was a "ground truth" and tell us limitations of The Cannon model (i.e. we could say, for input spectra with zero label uncertainty, The Cannon can reproduce flux at each pixel with less than $XX\%$ error (comment on how well The Cannon reconstructs important features), and can reproduce labels to $XX\%$ uncertainty). Compare this to Mann and other traiing sets. Not hard to add, but having some issues with continuum normalization.}\color{black}

@@

\item[-] Analysis of specific features? \color{red}{Can easily plot some line lists, but not sure where to go with the analysis... What features are important for M dwarfs? Compare derivative plots of specific lines between theory-trained models and data-trained models?}\color{black}

\item[-] Other assumptions / issues?
\end{enumerate}

%%==========================================================================================

\bibliographystyle{aasjournal}
\bibliography{ref}

>>>>

% \begin{enumerate}
% \item[-] Why M-dwarfs are important: most abundant type of star, are planet hosts (The CARMENES search for exoplanets around M dwarfs \citealt{Trifonov:2018}; some famous examples: Trappist 1 (M8) -- \citealt{Gillon:2017}; Kepler-186, Kepler-138... other examples?), have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of Milky way populations \citep{Bochanski:2007}.
% \end{enumerate}
Very low mass (VLM) are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and $\sim 40 \%$ by mass 
 are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of Milky way populations \citep{Bochanski:2007}.

\subsection{M dwarf Challenge}
\begin{enumerate}
\item[-] Various attempts / samples for which people have tried to make substantial catalogs of M-dwarfs with physical labels on them (\citealt{West:2011} is the largest M dwarf catalog of $\sim$ 70k sources; \citealt{Deacon:2014} -- \color{red}{Need citations!}\color{black})

\item[-] Why is there a problem labeling M dwarfs? Generally, two kinds of approaches to labeling:

 (1) model the spectra using atmospheric models (BTSettl, \citealt{Allard:2011}; PHOENIX, \citealt{Husser:2013}, MARCS \citealt{Gustafsson:2008}, ATLAS \citealt{Castelli:2004}, etc.) find some places in literature where people have tried these on M dwarfs (\color{red}{Need citations... }\color{black} look at Burgasser, Blake...) - however missing opacities, and complexities due to molecules/clouds/chemistry make this difficult (\color{red}{Need citations... }\color{black} \citealt{Allard:2013});

 (2) model narrow wavelegth region where models are good, look at specific features (\citealt{Rojas-Ayala:2012}...), or reduce resolution (\citealt{Casagrande:2008} estimated effective temperatures, bolometric luminosities, and metallicities using optical and infrared photometries) -- however we want to use \emph{all} of the data!

@@

\item[-] Describe M dwarf studies that have already been conducted in APOGEE (\citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs; \citealt{Souto:2017} -- modeled 2 M dwarfs, determined Teff/logg/metallicity + 13 elemental abundances; \citealt{Desphande:2013} -- modeled rotational and radial velocites for several hundred M dwarfs; \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity)

@@

A number of M dwarf studies

@@

\subsection{APOGEE Survey}
% \begin{enumerate}
% \item[-] Describe the APOGEE data (survey mission: \citealt{Majewski:2015}; target selection: \citealt{Zasowski:2013}; ASPCAP pipeline: \citealt{Perez:2016}).

% \item[-] Some M dwarfs in APOGEE proposed as ancilliary targets (Blake/Covey sources), plus an unknown number observed unintentionally \citep{Desphande:2013}.
% \end{enumerate}

The APOGEE survey \citep{Majewski:2015} of the Sloan Digital Sky Survey IV (SDSS-IV) is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m) survey which has observed over 260,000 stellar spectra since its fourteenth data release (DR14; \citealt{Abolfathi:2017}). While APOGEE is primarily targeted for bright stellar populations (descirbe cuts \citealt{Zasowski:2013}) with the objective of studying galactic composition and evolution, numerous cool, main sequence have also been observed either as targets proposed by the M dwarf ancilliary survey \citep{Desphande:2013}

Fundamental parameters for each of these stars (including temperature, surface gravity, metallicity and 15 chemical abundances) are estimated by the APOGEE Stellar Parameter
and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016})

@@

%%==========================================================================================
\section{Experimental Results} \label{sec:results}

\subsection{Spectral Type Model}

\begin{enumerate}
\item[-] We trained The Cannon on 51 M dwarfs with a one-dimensional spectral type label, and was able to achieve a precision of $\pm$0.8 spectral types, more precise than the original training label uncertainty of $\pm$1 spectral type.
\item[-] Two tests of consistency were used to assess the validity of our derived test labels: Figure \ref{fig:west_selftest} displays the label self-test for the spectral type model, and Figure 2 displays 
\end{enumerate}

\begin{figure}[ht]
\plotone{figures/west_spt_self_test.png}
\caption{Spectral type model: training label vs. test label.} \label{fig:west_selftest}
\end{figure}

\begin{figure}[ht]
\plotone{figures/1_Spectral_Sequence.pdf}
\plotone{figures/2_Spectral_Sequence.pdf}
\plotone{figures/3_Spectral_Sequence.pdf}
\caption{ Spectral sequence of dwarfs in training set M0-M9; chip 1 of APOGEE
spectrum with highlighted spectral type sensitive regions identified in \citealt{Desphande:2013}.} \label{fig:sp_sequence}
\end{figure}

\subsection{Temperature/Metallicity Model}
\begin{figure}[ht]
\plottwo{figures/mann_teff_self_test.png}{figures/mann_feh_self_test.png}
\caption{Temperature and Metallicity label self test.} \label{fig:mann_selftest}
\end{figure}

\begin{figure}[ht]
\plotone{figures/1_physical_model.png}
\plotone{figures/2_physical_model.png}
\plotone{figures/3_physical_model.png}
\caption{\textit{Top two panels:} Mann-trained model for varying temperatures; \textit{bottom two panels:} Mann-trained model for varying metallicities.} \label{fig:mann_demo}
\end{figure}

@@

@@

\item[-] Comment on specific features? \color{red}{Can easily plot some line lists, but not sure where to go with the analysis... What features are important for M dwarfs? Compare derivative plots of specific lines between theory-trained models and data-trained models?}\color{black}

\item[-] The Cannon as a test of evaluating training parameter accuracy? 
\end{enumerate}

\begin{figure}[ht]
\plotone{figures/west_spt_derivative.png}
\caption{Derivative plots for spectral type model.} \label{fig:west_derivative}
\end{figure}

\begin{figure}[ht]
\plotone{figures/mann_teff_derivative.png}
\plotone{figures/mann_feh_derivative.png}
\caption{Derivative plots for physical parameter model. \textit{Top:} Derivative of flux with repsect to temperature. \textit{Bottom:} Derivative of flux with repsect to metallicity.} \label{fig:mann_derivative}
\end{figure}

%%==========================================================================================
\clearpage
\bibliographystyle{aasjournal}
\bibliography{ref}

### 2018-03-26 paper/outline/outline.tex
add derivative plots

\begin{enumerate}
\item[-] Why M-dwarfs are important: most abundant type of star, are planet hosts (Trappist 1, other examples?), have lifespans longer than the age of the universe (billions or up to trillions of years) making them an interesting probe of Milky way populations. \color{red}{Need citations!}
\end{enumerate}

\subsection{M dwarf Challenge}
\begin{enumerate}
\item[-] Various attempts / samples for which people have tried to make substantial catalogs of M-dwarfs with physical labels on them (\citealt{West:2011} is the largest M dwarf catalog of $\sim$ 70k sources... other catalogs? -- \color{red}{Need citations!}\color{black})

\item[-] Why is there a problem labeling M dwarfs? Generally, two kinds of approaches to labeling: (1) model the spectra using atmospheric models (BTSettl, PHOENIX, KURUCZ, MARCS, ATLAS, etc.) find some places in literature where people have tried these on M dwarfs (\color{red}{Need citations... }\color{black} look at Burgasser, Blake...) - however missing opacities, and complexities due to molecules/clouds/chemistry make this difficult (\color{red}{Need citations... }\color{black} look at Allard, Homeier...); (2) model narrow wavelegth region where models are good, look at specific features (\citealt{Rojas-Ayala:2012}...), or reduce resolution (use photometries) -- however we want to use \emph{all} of the data!

@@

%%==========================================================================================

\bibliographystyle{aasjournal}
\bibliography{ref}

>>>>

\begin{enumerate}
\item[-] Why M-dwarfs are important: most abundant type of star, are planet hosts (The CARMENES search for exoplanets around M dwarfs \citealt{Trifonov:2018}; some famous examples: Trappist 1 (M8) -- \citealt{Gillon:2017}; Kepler-186, Kepler-138... other examples?), have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of Milky way populations \citep{Bochanski:2007}. \color{red}{Need citations!}
\end{enumerate}

\subsection{M dwarf Challenge}
\begin{enumerate}
\item[-] Various attempts / samples for which people have tried to make substantial catalogs of M-dwarfs with physical labels on them (\citealt{West:2011} is the largest M dwarf catalog of $\sim$ 70k sources; \citealt{Deacon:2014} -- \color{red}{Need citations!}\color{black})

\item[-] Why is there a problem labeling M dwarfs? Generally, two kinds of approaches to labeling:

 (1) model the spectra using atmospheric models (BTSettl, PHOENIX, KURUCZ, MARCS, ATLAS, etc.) find some places in literature where people have tried these on M dwarfs (\color{red}{Need citations... }\color{black} look at Burgasser, Blake...) - however missing opacities, and complexities due to molecules/clouds/chemistry make this difficult (\color{red}{Need citations... }\color{black} \citealt{Allard:2013});

 (2) model narrow wavelegth region where models are good, look at specific features (\citealt{Rojas-Ayala:2012}...), or reduce resolution (\citealt{Casagrande:2008} estimated effective temperatures, bolometric luminosities, and metallicities using optical and infrared photometries) -- however we want to use \emph{all} of the data!

@@

%%==========================================================================================
\section{Future Work} \label{sec:future}

\begin{enumerate}
\item[-] M dwarf pipeline: with The Cannon we can identify and label \emph{all} of the M dwarfs in APOGEE.
\item[-] Describe ideal Cannon input (for example using FGK pairs for very accurate metallicites).
\end{enumerate}

%%==========================================================================================
\section{Conclusion} \label{sec:conclusion}
Reiterate findings.

%%==========================================================================================
\clearpage
\bibliographystyle{aasjournal}
\bibliography{ref}
